Concept of a Double Tilted Rowland Spectrograph for X-rays

### 2024-03-25 papers/Doubletiltconcept/doubletiltconcept.tex
ApJ double tilt paper ready for submission

\begin{abstract}
High-resolution spectroscopy in soft X-rays ($<2$~keV) requires diffractive elements to resolve any astrophysically relevant diagnostics, such as closely spaced lines, weak absorption lines, or line profiles. The Rowland torus geometry describes how gratings and detectors need to be positioned to optimize the spectral resolving power. We describe how an on-axis Rowland geometry can be tilted to accommodate blazed gratings. In this geometry, two channels with separate optical axes can share the same detectors (double tilted Rowland spectrograph, DTRS). Small offsets between the channels can mitigate the effect of chip gaps and reduce the alignment requirements during the construction of the instrument. The DTRS concept is especially useful for sub-apertured mirrors, because it allows an effective use of space in the entrance aperture of a spacecraft.
One mission that applies this concept is the Arcus probe.

@@

In this article, we present an optical layout for an instrument that stacks two or four different channels with distinct optical axes in such a way that each channel follows the Rowland torus prescription, yet at the same time all channels are imaged onto a common set of detectors. The general concept could be applied to a large range of focal lengths, grating constants, and other hardware properties. Thus, we discuss advantages and challenges of this layout in general. However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA MIDEX (medium explorer) and probe calls, applies this layout. Specific details of the  Arcus performance, such as predicted effective area and resolving power are shown in \citet{2018SPIE10699E..6FG,2023SPIE12678E..1DG}, but the general concept of a double tilted Rowland spectrograph (DTRS) with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.

The structure of this article is as follows: First, we discuss the properties on the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}. Then, we illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS. In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept in general and for the example of Arcus in particular (section~\ref{sect:applications}). We end with a short summary in section \ref{sect:summary}.

@@

\subsection{Mirror characteristics and sub-aperturing}
The concept of a DTRS can be implemented with different mirror technologies. In X-rays, the mirror assembly will typically consist of nested (full or partial) shells that reflect X-rays in grazing incidence. The higher the energy of the X-ray, the smaller the angle between the mirror surface and the incoming ray has to be to allow for reflection. The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $\lambda / \Delta \lambda$ increases when the mirror point-spread-function (PSF) decreases. Depending on the type of mirror chosen, the PSF can be dominated by the alignment error between mirror shells, in particular for mirror assemblies that consist of hundreds of elements. In this case, X-rays are offset from the focus point both in the plane of reflection and perpendicular to it.
On the other hand, mirrors can also have a figure error, which is the deviation of the mirror surface from the ideal shape. If the PSF is dominated by the figure error or scattering by microroughness or particulates, then scattering angles in the plane of reflection are typically larger than the out-of-plane scatter. In this case, sampling a limited azimuthal span of the aperture -- subaperturing -- can improve the spectral resolving power \citep{1987ApOpt..26.2915C}. This is illustrated in Fig.~\ref{fig:subaperture}.

@@

\subsection{Dispersive elements}
In a Rowland torus geometry, diffraction gratings are placed on the surface of the Rowland torus, ideally tangentially for transmission gratings. They diffract light according to the grating equation for photons arriving perpendicular to the grating surface:
\begin{equation}
n \lambda = d \sin \theta \label{eqn:diffraction}
\end{equation}
where $\lambda$ is the wavelength of the light, $n$ is the order of the diffraction, $d$ is the grating constant, and $\theta$ is the angle between the incoming light and the normal to the grating. (The equation can be rewritten for rays hitting the grating at arbitrary angles, but for the purpose of this conceptual discussion this is not needed.)
Typically, these transmission gratings are flat and only a few cm$^2$ in size; the spacing between gratings bars $d$ is the same for all gratings and does not vary within a grating. While a flat grating necessarily deviates from the curved surface of a torus, in today's instruments those deviations are small enough that they do not impact the resolving power of the spectrograph significantly. Larger gratings would increase the effective area of the instrument because less space is needed for frames and other mounting structures that hold the gratings in place, but they would also deviate more from the surface of the torus. \citet{2020SPIE11444E..88G} explored bending the gratings in one dimension or varying the distance between grating bars over a grating to compensate for the deviations from the ideal shape in simulations. However, these concepts have not yet been implemented in X-ray transmission gratings.

@@

Figure~\ref{fig:sketch_label} shows sketches of the geometry in the plane spanned by the optical axis and the symmetry axis of the torus. Details will be discussed in the following subsections.
\begin{figure*}
    \includegraphics[width=0.5\textwidth]{sketch_label}
    \includegraphics[width=0.5\textwidth]{raytrace_labelled}
    \caption{\emph{left:} Sketch of a tilted Rowland torus geometry shown in the plane of the optical axis and the symmetry axis of the torus. Photons arrive from the left. The location of the mirror is shown as blue box. The optical axis (solid blue line) is offset from the center of the Rowland circle (blue square) and is perpendicular to the symmetry axis of the torus (blue long-dashed line). The two blue circles show the cut through the 3D torus in this plane; they are connected by blue dotted lines to help visualize the 3D position of the torus. $\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (A) to the center of the Rowland circle. $\beta$ is the angle from optical axis to the line connecting A to the ``hinge'' -- the other point in this plane where symmetry axis of the torus and the Rowland circle intersect.
    \emph{right:} Visualization of a ray-trace of the same geometry. Colored lines show the path of individual rays starting on the left. Rays detected in the zeroth order are shown in purple, green and yellow are higher orders. Optical axis, the axis of the Rowland torus etc.\ are indicated in blue and the location of detectors in orange, matching the sketch in the left panel.
        }
    \label{fig:sketch_label}
\end{figure*}
\begin{figure*}
    \plotone{sketch}
    \caption{\emph{left:} Sketch of the on-axis Rowland geometry in the plane of the optical axis and the symmetry axis of the torus. Photons arrive from the left.
    \emph{center:} Tilted Rowland torus. The optical axis of the blue channel is offset from the center of the Rowland circle, but intersects the symmetry axis of the torus in the focal point $F$.
    \emph{right:} Layout with two Rowland tori, tilted in opposite direction. In the plane spanned by both optical axes (which are parallel) and the symmetry axis of the torus, the two tori overlap in exactly one circle, the Rowland circle. See figure~\ref{fig:sketch_label} for a more detailed explanation.
        }
    \label{fig:sketch}
\end{figure*}

\subsection{One Rowland Torus with central optical axis}
\label{sect:onetorus}
For a single Rowland torus with no tilt, the optical axis passes through the center of the Rowland circle and $R=r$ as shown in the left panel of figure~\ref{fig:sketch}. Figure~\ref{fig:3d:single} shows a 3D view of a ray-trace. Since the diffraction gratings are symmetric with respect to the optical axis, positive or negative diffraction orders (up or down in figure~\ref{fig:sketch}) are equally likely, and the diffracted photons will be symmetric with respect to the focal point. To avoid the same wavelength to fall in a chip gap on the detector for both the positive and the negative diffraction orders, it is useful to position the CCDs with an offset, so that the chip gaps are not symmetric.
\begin{figure}
  \begin{interactive}{js}{ifigs/on-axis.zip}
  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_overview}
  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_aperture}
  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_gratings}
  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_detector}
  \end{interactive}
\caption{3D image of the ray-trace for a source with a continuum spectrum over a limited bandpass, seen from the side, from the top (where the rays enter the aperture), zoom on the grating positions, and zoom on the detector. Rays are shown from the entrance aperture on. They pass a focussing mirror represented by a green box and get diffracted by gratings (white squares). They are detected by CCDs (yellow). Rays are colored by the diffraction order: Light purple (middle) are zero order photons. A section of the Rowland torus is shown as transparent gray. The big square at the right indicates the position of the focal plane. Rays are traced past the detectors to the focal plane to show how the CCDs on the Rowland circle capture the rays where the dispersion width is smallest - past the CCDs the rays widen in dispersion direction again, but still become narrower in cross-dispersion direction. This figure is interactive in the online version, allowing the reader to pan, zoom, and rotate to see those details and inspect the rays from all angles. (Use the mouse to rotate, ``ctrl'' and mouse to pan, mouse wheel to zoom, and double-click to set the center of rotation. On other devices, other native controls might be available, e.g.\ two-finger touch to zoom on touchscreens.)
\label{fig:3d:single}}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.6\textwidth]{ondetector}
    \caption{Distribution of photons on the detectors for different designs. Photons are colored by channel and the position of the zero(s) order(s) is highlighted with red boxes. Note that the dispersion and cross-dispersion directions are not to scale. In the lower two panels the zeroths orders of one channel overlap with the dispersed phontons from another channel. This could be avoided with larger offsets between channels in cross-dispersion direction. Also, in real instruments with focal lengths significantly larger than in the illustrative model we use throughout this article, the bending and the width in cross-dispersion direction are typically smaller, so that this is not a problem in practice.`'
        }
    \label{fig:fish}
\end{figure}

\citet{doi:10.1117/12.856482} suggested tilting the torus with respect to the optical axis when using critical angle transmission (CAT) gratings \citep{2021hai4.book..235H,2022ApJ...934..171H}. In CAT gratings, the grating bars have a high aspect ratio, that is the distance between two gratings bars is much smaller than the height of the bars. CAT gratings are typically used at a blaze angle, where the grating bars are inclined with respect to the incoming rays. CAT gratings preferentially diffract photons to twice the blaze angle and can reach high diffraction efficiencies. However, the blaze angle also means that the gratings can deviate significantly from the surface of the  Rowland torus. To minimize this deviation, the Rowland torus can be tilted.
This tilt can be described by two angles $\alpha$ and $\beta$ as shown in figure~\ref{fig:sketch_label}. $\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (labelled ``A'' in the figure) and the center of the Rowland circle.  We choose the distance between point A and the focal point F to be slightly shorter than the focal length $f$ of the mirror to allow for space to mount the gratings behind the mirror and use the symbol ``a'' for the distance between A and F.
In this configuration, the torus has $R < r$ so there are two points where the torus overlaps itself in the plane spanned by the optical axis and the symmetry axis of the torus (the plane shown in figure~\ref{fig:sketch_label}). The first of those intersection points is F. We call the second point ``Hinge''.
$\beta$ is the angle from optical axis to the line connecting A and the Hinge. From this we can derive the radius of the Rowland circle as
$$r = \frac{a}{2\cos \alpha}.$$
 If the focal point F is the origin of the coordinate system, then the center of the Rowland circle (blue square in left blue circle in the panels in figure~\ref{fig:sketch_label}) is at location
 \begin{equation}
    \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
    \begin{pmatrix} \frac{a}{2} \tan\alpha \\0 \\   \frac{a}{2} \end{pmatrix}.
 \end{equation}

@@

Assuming CAT gratings where the gratings bars are perpendicular to the surface of the grating, the gratings need to be installed blazed (tilted) with respect to the incoming rays. Thus, they deviate more from the Rowland-torus than a grating that is placed perpendicular to the incoming rays. Figure~\ref{fig:3d:tilted} shows this layout in 3D, but with a very large blaze angle for clarity; for real CAT gratings typical blaze angles are 1-2~deg. Tilting the Rowland torus by about twice the blaze angle reduces the average deviation of the grating bar position from te surface of the Rowland torus. Also, CAT gratings disperse predominantly in one direction. However, the Rowland geometry optimizes the spectral resolving power by minimizing the spread of the photons in dispersion direction; detectors are located on the Rowland circle, which represents the position of minimal spread in dispersion direction. This is not the same location as the imaging focal plane, which is defined by minimizing the total size of the PSF. At the focal point, the Rowland circle and the imaging focal plane match. Figure~\ref{fig:fish} shows how the dispersed photons spread out in cross-dispersion direction with increasing distance from F, roughly resembling the outline of a fish. With a tilted Rowland torus, there is a second intersection point between the spectral and the imaging focus where the cross-dispersion width becomes small (the point between the body and the tail of the fish, the ``Hinge'', about +300~mm in dispersion direction in the second panel of figure~\ref{fig:fish}). This location can be chosen to be roughly in the center of the dispersed spectrum and thus limit the cross-dispersion profile to reduce the area of the detector that contributes to the extracted signal and thus the background. We note that the width in cross-dispersion direction is much larger for larger diffraction angles and, in real instruments, will be smaller than shown in the figure.

@@

\subsection{Advantages of the DTRT concept}
\subsubsection{Subaperturing space considerations, and resolving power}
Manufacturing mirrors with a smaller PSF is more difficult and costly than manufacturing mirrors with a larger PSF but otherwise identical properties. For a spectrograph, the width of the PSF in dispersion direction is far more important than the width in cross-dispersion direction. If the width in dispersion direction can be reduced by sub-aperturing, that provides a higher spectral resolving power at lower cost and technical complexity compared to reducing the PSF of the mirror itself. On the other hand, it reduces the collecting area of the mirror, since only a fraction of the full circle can be used. The DTRS design allows us to compensate that by having multiple channels, ultimately delivering better resolving power at the same mirror requirements than a single-channel design.

@@

\subsubsection{Zeroth order is seen}
Observing the direct image of all channels on the detector has several advantages. First, most of the high-energy photons are not diffracted by typical CAT gratings. So, the direct image can be used for analysis of high-energy spectra using the intrinsic energy resolution of the detector, e.g.\ a CCD or even microcalorimeter. For example, the Fe line at 6.7~keV is a widely used diagnostic for many astrophysical objects; in a DTRT design this line can be observed with an energy sensitive detector such as a CCD in the zeroth order, while the lower-energy photons are observed in the dispersed orders at the same time. Second, accurate knowledge of the position of the direct image is important for the wavelength calibration of the dispersed signal. The wavelength of the light is determined by measuring the distance $x$ from the detected photon to the zeroth order and the known distance $a$ between the gratings and the focal plane. Conceptually, $\tan \theta = x/a$. With the known grating constant $d$, equation~\ref{eqn:diffraction} then yields $\lambda$. For high spectral resolving power and large values of $\theta$ it might be important to take the 3d shape of the detector into account, but either way, accurate knowledge of the zero order position is crucial.

\subsubsection{Robustness}
The DTRS concept increases the robustness and reduces the technical risk of a mission in several ways. Because we can assign each sub-apertured mirror wedge its own channel (in the 4 channel design with offsets), there is no need to align the mirrors for each channel to each other to better than a cm or so.
Also, the design is robust against loosing individual detectors in a camera. Since half of the photons are dispersed from left-to-right and the others right-to-left, loss of a detector element in a camera does not result in a complete loss of spectral coverage for any wavelength. The multichannel design can also help to uncover systematics in the calibration.

@@

\section{Applications}
\label{sect:applications}
As a concrete example, we describe the Arcus X-ray spectrograph  \citep{2023SPIE12678E..0ES}. Arcus has gone through multiple iterations for different mission calls, most recently as a NASA X-ray Probe concept submitted in 2023. This mission adopts a DTRS design with $\alpha=3.6$~deg and $\beta=2\alpha$. It uses silicon-pore optics (SPO) with a 12~m focal length that are developed for the Athena mission \citep{2023SPIE12679E..05G}; $a$ is slightly smaller than the focal length with $a=11876$~mm. SPOs are manufactured as small stacks of about 30 mirror plates with sizes of order 10-20~cm in each direction (the exact size depends on the radius). 40 SPOs are aligned into one petal and each petal defines one of four optical channels in a DTRS design with offsets. The two sub-channels are offset by $\pm2.5$~mm in dispersion direction with respect to the design with just two Rowland tori. In cross-dispersion the four channels are located at $-7.5$, $-2.5$, $+2.5$, and $+7.5$~mm from the center of the CCDs.
CAT gratings are manufactured through deep reactive-ion etching of silicon-on-insulator wafers \citep{2022ApJ...934..171H,2023SPIE12679E..0LH}. These gratings are mounted with a blaze angle, i.e.\ they are intentionally tilted by 1.8~deg with respect to the incoming photons, which concentrates the diffracted photons around the ``blaze peak'' at a diffraction angle of $2 * 1.8 = 3.6$~deg, so that only 8~CCDs are needed in each camera covering about 30~cm in the dispersion direction with a gap close to 50~cm between the two cameras. Arcus will achieve a spectral resolving power around 3500 for wavelength between 1.5 and 6.0~nm, averaged over all diffraction orders, and an effective area up to 500~cm$^2$ between 1.5 and 2.5~nm, dropping to 100~cm$^2$ at 6.0~nm. Detailed ray-tracing simulations for Arcus that optimize the parameters of the design and predict the performance depending on the misalignment for the different components are shown in \citet{2017SPIE10397E..0PG,2018SPIE10699E..6FG,2023SPIE12678E..1DG}.

>>>>

\begin{abstract}
High-resolution spectroscopy in soft X-rays ($<2$~keV) requires diffractive elements to resolve any astrophysically relevant diagnostics, such as closely spaced lines, weak absorption lines, or line profiles. The Rowland torus geometry describes how gratings and detectors need to be positioned to optimize the spectral resolving power. We describe how an on-axis Rowland geometry can be tilted to accommodate blazed gratings. In this geometry, two channels with separate optical axes can share the same detectors (double tilted Rowland spectrograph, DTRS). Small offsets between the channels can mitigate the effect of chip gaps and reduce the alignment requirements during the construction of the instrument. The DTRS concept is especially useful for sub-apertured mirrors, because it allows an effective use of space in the entrance aperture of a spacecraft.
One mission that applies this concept is the Arcus Probe.

@@

In this article, we present an optical layout for an instrument that stacks two or four different channels with distinct optical axes in such a way that each channel follows the Rowland torus prescription, yet at the same time all channels are imaged onto a common set of detectors. The general concept could be applied to a large range of focal lengths, grating constants, and other hardware properties. Thus, we discuss advantages and challenges of this layout in general. However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA MIDEX (medium explorer) and Probe calls, applies this layout. Specific details of the  Arcus performance, such as predicted effective area and resolving power are shown in \citet{2018SPIE10699E..6FG,2023SPIE12678E..1DG}, but the general concept of a double tilted Rowland spectrograph (DTRS) with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.

The structure of this article is as follows: First, we discuss the properties of the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}. Then, we illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS. In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept in general and for the example of Arcus in particular (section~\ref{sect:applications}). We end with a short summary in section \ref{sect:summary}.

@@

\subsection{Mirror characteristics and sub-aperturing}
The concept of a DTRS can be implemented with different mirror technologies. In X-rays, the mirror assembly will typically consist of nested (full or partial) shells that reflect X-rays in grazing incidence. The higher the energy of the X-ray, the smaller the angle between the mirror surface and the incoming ray has to be to allow for efficient reflection. The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $\lambda / \Delta \lambda$ increases when the mirror point-spread-function (PSF) decreases. Depending on the type of mirror chosen, the PSF can be dominated by the alignment error between mirror shells, in particular for mirror assemblies that consist of hundreds of elements. In this case, X-rays are offset from the focus point both in the plane of reflection and perpendicular to it.
On the other hand, mirrors can also have a figure error, which is the deviation of the mirror surface from the ideal shape. If the PSF is dominated by the figure error or scattering by microroughness or particulates, then scattering angles in the plane of reflection are typically larger than the out-of-plane scatter. In this case, sampling a limited azimuthal span of the aperture -- subaperturing -- can improve the spectral resolving power \citep{1987ApOpt..26.2915C,2010SPIE.7732E..1JH}. This is illustrated in Fig.~\ref{fig:subaperture}.

@@

\subsection{Dispersive elements}
In a Rowland torus geometry, diffraction gratings are placed on the surface of the Rowland torus, ideally tangentially for transmission gratings. They diffract light according to the grating equation for photons arriving perpendicular to the grating surface:
\begin{equation}
n \lambda = d \sin \theta \label{eqn:diffraction}
\end{equation}
where $\lambda$ is the wavelength of the light, $n$ is the order of the diffraction, $d$ is the grating constant, and $\theta$ is the angle between the incoming light and the normal to the grating. (The equation can be rewritten for rays hitting the grating at arbitrary angles, but for the purpose of this conceptual discussion this is not needed.)
Typically, these transmission gratings are flat and only a few cm$^2$ in size; the spacing between gratings bars $d$ is the same for all gratings and does not vary within a grating. While a flat grating necessarily deviates from the curved surface of a torus, in today's instruments those deviations are small enough that they do not impact the resolving power of the spectrograph significantly. Larger gratings would increase the effective area of the instrument because less space is needed for frames and other mounting structures that hold the gratings in place, but they would also deviate more from the surface of the torus. \citet{2020SPIE11444E..88G} explored bending the gratings in one dimension or varying the distance between grating bars over a grating to compensate for the deviations from the ideal shape in simulations. \citet{2019SPIE11118E..11G} shows that CAT gratings can be bent to the required radii without degrading their performance.

@@

Figure~\ref{fig:sketch_label} shows sketches of the geometry in the plane spanned by the optical axis and the symmetry axis of the torus. Details will be discussed in the following subsections.
\begin{figure*}
    \includegraphics[width=0.5\textwidth]{sketch_label}
    \includegraphics[width=0.5\textwidth]{raytrace_labelled}
    \caption{\emph{left:} Sketch of a tilted Rowland torus geometry shown in the plane of the optical axis and the symmetry axis of the torus. Photons arrive from the left. The location of the mirror is shown as blue box. The optical axis (solid blue line) is offset from the center of the Rowland circle (blue square). The symmetry axis of the torus  is marked as blue long-dashed line. The two blue circles show the cut through the 3D torus in this plane; they are connected by blue dotted lines to help visualize the 3D position of the torus. $\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (A) to the center of the Rowland circle. $\beta$ is the angle from optical axis to the line connecting A to the ``hinge'' -- the other point in this plane where symmetry axis of the torus and the Rowland circle intersect.
    \emph{right:} Visualization of a ray-trace of the same geometry. Colored lines show the path of individual rays starting on the left. Rays detected in the zeroth order are shown in purple, green and yellow are higher orders. Optical axis, the axis of the Rowland torus etc.\ are indicated in blue and the location of detectors in orange, matching the sketch in the left panel.
        }
    \label{fig:sketch_label}
\end{figure*}
\begin{figure*}
    \plotone{sketch}
    \caption{\emph{left:} Sketch of the on-axis Rowland geometry in the plane of the optical axis and the symmetry axis of the torus. Photons arrive from the left.
    \emph{center:} Tilted Rowland torus. The optical axis of the blue channel is offset from the center of the Rowland circle, but intersects the symmetry axis of the torus in the focal point $F$.
    \emph{right:} Layout with two Rowland tori, tilted in opposite direction. In the plane spanned by both optical axes (which are parallel) and the symmetry axis of the tori, the two tori overlap in exactly one circle, the Rowland circle. See figure~\ref{fig:sketch_label} for a more detailed explanation.
        }
    \label{fig:sketch}
\end{figure*}

\subsection{One Rowland Torus with central optical axis}
\label{sect:onetorus}
For a single Rowland torus with no tilt, the optical axis passes through the center of the Rowland circle and $R=r$ as shown in the left panel of figure~\ref{fig:sketch}. Figure~\ref{fig:3d:single} shows a 3D view of a ray-trace. Since the diffraction gratings are symmetric with respect to the optical axis, positive or negative diffraction orders (up or down in figure~\ref{fig:sketch}) are equally likely, and the diffracted photons will be symmetric with respect to the focal point. To avoid the same wavelength to fall in a chip gap on the detector for both the positive and the negative diffraction orders, it is useful to position the CCDs with an offset, so that the chip gaps are not symmetric.
\begin{figure}
  \begin{interactive}{js}{ifigs/on-axis.zip}
  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_overview}
  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_aperture}
  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_gratings}
  \includegraphics[width=0.22\textwidth]{ifigs/on-axis_detector}
  \end{interactive}
\caption{3D images of the ray-trace for a source with a continuum spectrum over a limited bandpass, seen (from left to right) from the side, from the top (where the rays enter the aperture), zoom on the grating positions, and zoom on the detector. Rays are shown from the entrance aperture on. They pass a focussing mirror represented by a green box and get diffracted by gratings (white squares). They are detected by CCDs (yellow). Rays are colored by the diffraction order: Light purple are zero order photons. A section of the Rowland torus is shown as transparent gray. The big square at the right indicates the position of the focal plane. Rays are traced past the detectors to the focal plane to show how the CCDs on the Rowland circle capture the rays where the dispersion width is smallest - past the CCDs the rays widen in dispersion direction again, but still become narrower in cross-dispersion direction. This figure is interactive in the online version, allowing the reader to pan, zoom, and rotate to see those details and inspect the rays from all angles. (Use the mouse to rotate, ``ctrl'' and mouse to pan, mouse wheel to zoom, and double-click to set the center of rotation. On other devices, other native controls might be available, e.g.\ two-finger touch to zoom on touchscreens.)
\label{fig:3d:single}}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.6\textwidth]{ondetector}
    \caption{Distribution of photons on the detectors for different designs. Photons are colored by channel and the position of the zero(s) order(s) is highlighted with red boxes. Note that the dispersion and cross-dispersion directions are not to scale. In the lower two panels the zeroths orders of one channel overlap with the dispersed phontons from another channel. This could be avoided with larger offsets between channels in cross-dispersion direction. Also, in real instruments with focal lengths significantly larger than in the illustrative model we use throughout this article, the bending and the width in cross-dispersion direction are typically smaller, so that this is not a problem in practice.
        }
    \label{fig:fish}
\end{figure}

\citet{doi:10.1117/12.856482} suggested tilting the torus with respect to the optical axis when using critical-angle transmission (CAT) gratings \citep{2021hai4.book..235H,2022ApJ...934..171H}. In CAT gratings, the grating bars have a high aspect ratio, that is the distance between two gratings bars is much smaller than the height of the bars. CAT gratings are typically used at a blaze angle, where the grating bars are inclined with respect to the incoming rays. CAT gratings preferentially diffract photons to twice the blaze angle and can reach high diffraction efficiencies. However, the blaze angle also means that the gratings can deviate significantly from the surface of the  Rowland torus. To minimize this deviation, the Rowland torus can be tilted.
This tilt can be described by two angles $\alpha$ and $\beta$ as shown in figure~\ref{fig:sketch_label}. $\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (labelled ``A'' in the figure) and the center of the Rowland circle.  We choose the distance between point A and the focal point F to be slightly shorter than the focal length $f$ of the mirror to allow for space to mount the gratings behind the mirror and use the symbol ``a'' for the distance between A and F.
In this configuration, the torus has $R < r$ so there are two points where the torus overlaps itself in the plane spanned by the optical axis and the symmetry axis of the torus (the plane shown in figure~\ref{fig:sketch_label}). The first of those intersection points is F. We call the second point ``Hinge''.
$\beta$ is the angle from optical axis to the line connecting A and the Hinge. From this we can derive the radius of the Rowland circle as
$$r = \frac{a}{2\cos \alpha}.$$
 If the focal point F is the origin of the coordinate system, then the center of the Rowland circle (blue square in left blue circle in the panels in figure~\ref{fig:sketch_label}) is at location
 \begin{equation}
    \begin{pmatrix} x \\ y \\ z \end{pmatrix} =
    \begin{pmatrix} \frac{a}{2} \tan\alpha \\0 \\   \frac{a}{2} \end{pmatrix}.
 \end{equation}

@@

Assuming CAT gratings where the gratings bars are perpendicular to the surface of the grating, the gratings need to be installed blazed (tilted) with respect to the incoming rays. Thus, they deviate more from the Rowland-torus than a grating that is placed perpendicular to the incoming rays. Figure~\ref{fig:3d:tilted} shows this layout in 3D, but with a very large blaze angle for clarity; for real CAT gratings typical blaze angles are 1-2~deg. Tilting the Rowland torus by about twice the blaze angle reduces the average deviation of the grating bar position from the surface of the Rowland torus. Also, CAT gratings disperse predominantly in one direction. However, the Rowland geometry optimizes the spectral resolving power by minimizing the spread of the photons in dispersion direction; detectors are located on the Rowland circle, which represents the position of minimal spread in dispersion direction. This is not the same location as the imaging focal plane, which is defined by minimizing the total size of the PSF. At the focal point, the Rowland circle and the imaging focal plane match. Figure~\ref{fig:fish} shows how the dispersed photons spread out in cross-dispersion direction with increasing distance from F, roughly resembling the outline of a fish. With a tilted Rowland torus, there is a second intersection point between the spectral and the imaging focus where the cross-dispersion width becomes small (the point between the body and the tail of the fish, the ``Hinge'', about +300~mm in dispersion direction in the second panel of figure~\ref{fig:fish}). This location can be chosen to be roughly in the center of the dispersed spectrum and thus limit the cross-dispersion profile to reduce the area of the detector that contributes to the extracted signal and thus the background. We note that the width in cross-dispersion direction is much larger for larger diffraction angles and, in real instruments, will be smaller than shown in the figure.

@@

\subsection{Advantages of the DTRT concept}
\subsubsection{Subaperturing space considerations, and resolving power}
Manufacturing mirrors with a smaller PSF is more difficult and costly than manufacturing mirrors with a larger PSF but otherwise identical properties. For a spectrograph, the width of the PSF in dispersion direction is far more important than the width in cross-dispersion direction. If the width in dispersion direction can be reduced by sub-aperturing, a higher spectral resolving power is achieved at lower cost and technical complexity compared to reducing the PSF of the mirror itself. On the other hand, it reduces the collecting area of the mirror, since only a fraction of the full circle can be used. The DTRS design allows us to compensate for this by having multiple channels, ultimately delivering better resolving power at the same mirror requirements than a single-channel design.

@@

\subsubsection{Zeroth order is seen}
Observing the direct image of all channels on the detector has several advantages. First, most of the high-energy photons are neither diffracted nor absorbed by typical CAT gratings. So, the direct image can be used for analysis of high-energy spectra using the intrinsic energy resolution of the detector, e.g.\ a CCD or even microcalorimeter. For example, the Fe line at 6.7~keV is a widely used diagnostic for many astrophysical objects; in a DTRT design this line can be observed with an energy sensitive detector such as a CCD in the zeroth order, while the lower-energy photons are observed in the dispersed orders at the same time. Second, accurate knowledge of the position of the direct image is important for the wavelength calibration of the dispersed signal. The wavelength of the light is determined by measuring the distance $x$ from the detected photon to the zeroth order and the known distance $a$ between the gratings and the focal plane. Conceptually, $\tan \theta \approx x/a$. With the known grating constant $d$, equation~\ref{eqn:diffraction} then yields $\lambda$. For high spectral resolving power and large values of $\theta$ it might be important to take the 3d shape of the detector into account, but either way, accurate knowledge of the zero order position is crucial.

\subsubsection{Robustness}
The DTRS concept increases the robustness and reduces the technical risk of a mission in several ways. Because we can assign each sub-apertured mirror wedge its own channel (in the 4 channel design with offsets), there is no need to align the mirrors for each channel to the other channel to better than a cm or so.
Also, the design is robust against loosing individual detectors in a camera. Since half of the photons are dispersed from left-to-right and the others right-to-left, loss of a detector element in a camera does not result in a complete loss of spectral coverage for any wavelength. The multichannel design can also help to uncover systematics in the calibration.

@@

\section{Applications}
\label{sect:applications}
As a concrete example, we describe the Arcus X-ray spectrograph  \citep{2023SPIE12678E..0ES}. Arcus has gone through multiple iterations for different mission calls, most recently as a NASA X-ray Probe concept submitted in 2023. This mission adopts a DTRS design with $\alpha=3.6$~deg and $\beta=2\alpha$. It uses silicon-pore optics (SPO) with a 12~m focal length that are developed for the Athena mission \citep{2023SPIE12679E..05G}; $a$ is slightly smaller than the focal length with $a=11876$~mm. SPOs are manufactured as small stacks of about 30 mirror plates with sizes of order 10-20~cm in each direction (the exact size depends on the radius). 40 SPOs are aligned into one petal and each petal defines one of four optical channels in a DTRS design with offsets. The two sub-channels are offset by $\pm2.5$~mm in dispersion direction with respect to the design with just two Rowland tori. In cross-dispersion the four channels are located at $-7.5$, $-2.5$, $+2.5$, and $+7.5$~mm from the center of the CCDs.
CAT gratings are manufactured through deep reactive-ion etching of silicon-on-insulator wafers \citep{2022ApJ...934..171H,2023SPIE12679E..0LH}. These gratings are mounted with a blaze angle, i.e.\ they are intentionally tilted by 1.8~deg with respect to the incoming photons, which concentrates the diffracted photons around the ``blaze peak'' at a diffraction angle of $2 * 1.8 = 3.6$~deg, so that only 8~CCDs are needed in each camera covering about 30~cm in the dispersion direction with a gap close to 50~cm between the two cameras. Arcus will achieve a spectral resolving power around 3500 for wavelengths between 1.5 and 6.0~nm, averaged over all diffraction orders, and an effective area up to 500~cm$^2$ between 1.5 and 2.5~nm, dropping to 100~cm$^2$ at 6.0~nm. Detailed ray-tracing simulations for Arcus that optimize the parameters of the design and predict the performance depending on the misalignment for the different components are shown in \citet{2017SPIE10397E..0PG,2018SPIE10699E..6FG,2023SPIE12678E..1DG}.
